On the propriety of restricted Boltzmann machines

### 2016-10-19 writing/draft.Rmd
through edits on uninterpretability

### Uninterpretability
@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by marginal mean-structure not being maintained in the model. A measure of this is the magnitude of the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$. Using this concept it is possible to investigate what conditions lead to uninterpretability in a model versus those that guarantee interpretable models. A model is considered without uninterpretability if its expected value is not very different from the same model with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}
and
\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol 
theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
&= \frac{\sum\limits_{\boldsymbol x \in \mathcal{X}} x_k \exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}  \qquad k = 1, \dots, H + V
\end{align*}

If $|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \emptyset\right]$ is large then the RBM with parameter vector $\boldsymbol \theta$ $is uninterpetable.

>>>>

### Uninterpretability
@kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by a marginal mean-structure failing to hold when dependence parameters in a model become too extreme. Note that, in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_V}$ and $\theta_{h_1}, \dots, \theta_{h_H}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_V, H_1, \dots, H_V$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, V, j = 1, \dots, H$. Hence, the main effect parameters have a clear interpretation under an independence model ($\theta_{ij} = 0$) but this interpretation (as well as an understanding of the nture of dependence) can break down as interaction parameters increase in magnitude (e.g. $\max |\theta_{i,j}| \rightarrow \infty$) relative to the size of the main effects. To access which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, at $\boldsymbol \theta$ and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$, where $\boldsymbol \theta^*$ is defined to equal $\boldsymbol \theta$ with all $\theta_{ij} = 0$ for $i = 1, \dots, V, j = 1, \dots, H$. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. Uninterpretability is avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation with assumed independence. The quantities to compare in the RBM case are

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x f_{\boldsymbol \theta}(\boldsymbol x) \\
&= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V \sum\limits_{j=1}^H \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}
\end{align*}
and
\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol 
\theta^*\right] &= \sum\limits_{\boldsymbol x \in \mathcal{X}} \boldsymbol x \frac{\exp\left(\sum\limits_{i = 1}^V \theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{X}}\exp\left(\sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^H\theta_{h_j} h_j\right)} \\
\end{align*}

If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable.
